Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-03-15 TEBenchmarkPaper/article.tex
Keep tables in place
Final fixes

% Derived from the template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{csvsimple}
\usepackage{hhline}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows, calc, decorations.pathreplacing, positioning,
  shapes.geometric}
% etc.

@@

One quantitative approach which has gained traction is \emph{precision/recall
  analysis}. This has been applied to IsaCoSy, IsaScheme and HipSpec, the latter
providing a direct comparison of these
tools~\cite{claessen2013automating}.\footnote{Unfortunately this comparison only
  reports prior results from the other tools, rather than reproducing them.
  This makes the numbers less comparable, since each tool was tested with
  slightly different definitions (e.g. whether or not the exponential and
  maximum functions were included).} To measure precision and recall we first
choose a set of definitions, such as those in figure \ref{figure:list_theory},
and a \emph{ground truth} to treat as the 'ideal' set of conjectures we would
like a tool to produce for those definitions. Each tool is run and their outputs
are compared to the ground truth. To score 100\% on precision and recall, all of
the conjectures which appear in the ground truth must be generated, and nothing
else:

@@

We can summarise runtimes by choosing their median, as this is more robust
against long-running outliers, and hence represents performance for a
``typical'' input of that size. Unlike previous one-off evaluations, we can also
compute the \emph{spread} of our data, for example with the inter-quartile
range.

The precision and recall for multiple samples can be combined in two ways. We
can calculate their mean value, known as the \emph{average of the ratios}, to
find the expected precision and recall for runs of this size. Alternatively, we
can sum their numerators and denominators to get the \emph{ratio of the
  averages}: this normalises the size of the generated and ground truth sets,
letting us model precision and recall as simple Bernoulli processes.

@@

We propose that running times are compared using a Wilcoxon signed-rank
test~\cite{wilcoxon1945individual}, since it does not require that the data be
normally distributed (unlike, say, the z-test or t-test). In our application we
follow a variant of the Speedup-Test protocol of Touati, Worms and
Briais~\cite{touati2013speedup}; the differences are our use of the paired form
of the Wilcoxon test, and the fact that each run was performed on a different
sample (to reduce sampling bias, at the expense of accuracy).

@@

We cannot use McNemar's test to compare precision, since each tool may generate
different conjectures. Instead, we use Boschloo's form of Barnard's test to
compare the number of wanted and unwanted conjectures generated by each tool.

@@

We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
benchmark as our ground truth corpus.

@@

Based on this we decided to benchmark sample sizes up to 20, since neither tool
seemed to perform well beyond that. The Speedup-test protocol follows the
statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
so we pick 31 samples of each size in order to cross this (arbitrary)
threshold. This gave a total of 1240 runs. To keep the benchmarking time down to
a few days we chose a timeout of 300 seconds, since that covered most of the
successful QuickSpec and IsaCoSy results we saw, and longer times gave rapidly
diminishing returns. During analysis, duplicate samples (caused by our sampling
restrictions) were found and discarded, so only 14 samples of size 1 were used
and 30 of size 2.

@@

We chose TIP for our ground truth since it satisfies the desiderata of
section~\ref{section:prep}: each problem has standalone type and function
definitions, making their separation trivial; known examples from the software
verification and inductive theorem proving literature are included, ensuring
relevance to those fields; the format includes higher-order functions and
inductive datatypes, which we are interested in; plus it is accompanied by
tooling to convert this format into a variety of languages, including Haskell
and Isabelle which are used by existing MTE tools.

@@

\begin{verbatim}
(declare-datatypes
  (a)                       ;; Type variable (element type)
  ((List                    ;; Type name
     (Nil)                  ;; Constructor (nullary)
     (Cons                  ;; Constructor (binary)
       (head a)             ;; Field name and type
       (tail (List a))))))  ;; Field name and type
\end{verbatim}

@@

\begin{verbatim}
(define-fun
  (par (a)                   ;; Type variable
    (constructor-Cons        ;; Function name
      ((x a) (xs (List a)))  ;; Argument names and types
      (List a)               ;; Return type
      (as                    ;; Type annotation
        (Cons x xs)          ;; Return value
        (List a)))))         ;; Return type

(define-fun
  (par (a)                        ;; Type variable
    (destructor-head              ;; Function name
      ((xs (List a)))             ;; Argument name and type
      a                           ;; Return type
      (match xs                   ;; Pattern-match
        (case (Cons h t) h)))))   ;; Return relevant field
\end{verbatim}

@@

Another complication is TIP's ``native'' support for booleans and integers,
which allows numerals and symbols like \texttt{+} to appear without any
accompanying definition. To ensure consistency in the translations, we replace
all occurrences of such expressions with standard definitions written with the
``user-level'' \texttt{declare-datatypes} and \texttt{define-fun}
mechanisms.~\footnote{\texttt{Boolean} has \texttt{true} and \texttt{false}
  constructors; \texttt{Natural} has \texttt{zero} and \texttt{successor};
  \texttt{Integer} has unary \texttt{positive} and \texttt{negative}
  constructors taking \texttt{Natural}s, and a nullary \texttt{zero} for
  symmetery.}

@@

TIP comes with software to translate its definitions into Haskell and Isabelle
code, including comparison functions and random data generators suitable for
QuickCheck. We translate all 269 unique definitions into a single module/theory
which is imported on each run of the tools, although only those functions which
appear in the current sample are actually explored.

@@

\begin{itemize}
\item We must decide what variables to include. We choose to add three variables
  for each type that appears as a function argument, except for types which have
  no QuickCheck data generators.
\item We must \emph{monomorphise} all types. For example, functions like
  \texttt{constructor-Cons} are \emph{polymorphic}: they build lists of any
  element type, but we need to pick a specific type in order to know which
  random value generator to use. We resolve this (arbitrarily) by picking
  \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
    \texttt{*} (types); for kind \texttt{* -> *} (type constructors) we pick
    \texttt{[]} (Haskell's list type constructor). If these violate some
    type class constraint, we pick a suitable type non-deterministically from
    those in scope during compilation; if no suitable type is found, we give up
    and don't include the  variable.}
\item Haskell functions are ``black boxes'', which QuickSpec can't compare
  during its exploration process. They are also curried, always taking one
  argument but potentially returning another function. QuickSpec lets us assign
  an arity to each function in the signature, from 0 to 5, so we pick the
  highest that is type-correct, since this avoids a proliferation of
  incomparable, partially-applied functions.
\end{itemize}

\begin{figure}
  \centering
  \input{quickspectime.pgf}
  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
    is an individual run, with 30 per size except for 2 (29). Colour indicates
    the total number of conjectures produced by each run, with failing processes
    shown in red. Points are spread out horizontally to prevent overlaps.
    Box plots show median and quartiles, with whiskers showing points at
    1.5$\times$~inter-quartile range.}
  \label{figure:quickspec_runtimes}
\end{figure}

@@

The time taken to explore samples of different sizes is shown in
\ref{figure:quickspec_runtimes}. Runs which timed out are shown in red, and
these occurred for all sample sizes but are more common for larger samples (over
half the runs for sample size 20 timed out). Runs which succeeded mostly
finished within 30 seconds, generating few or no conjectures; those taking
longer generated more conjectures, with the most being 116 conjectures from a
sample of size 16.

We model precision and recall as simple Bernoulli processes: the generated
conjectures (wanted or not) are like coin tosses whose bias is the precision;
likewise, the ground truth theorems (found or not) are like coin tosses biased
by the recall.

QuickSpec's precision gets worse as the sample size grows. Since QuickSpec
generates monotonically more conjectures as definitions are added to a signature
(assuming sufficient testing), it can't be finding fewer correct conjectures at
larger sizes. This is supported by the relative flatness of the recall results.
Rather, the number of conjectures generated is increasing at a higher rate than
the size of the ground truth. These extra conjectures may involve the
``padding'' definitions in a sample which don't contribute to its ground truth,
or may be ``uninteresting'' relationships between the dependencies of different
ground truth theorems.

This indicates two potential improvements to the QuickSpec algorithm (as far as
this benchmark is concerned). The deluge of generated conjectures could be
filtered by another post-processing step down to a more desirable sub-set.
Alternatively, large signatures could be filtered to remove ``padding''
definitions (those whose conjectures are unlikely to be ``interesting'') and
divided up into multiple smaller signatures to separate those definitions whose
relationships are ``uninteresting''. Both of these methods could improve the
precision, although care would be needed to avoid a large reduction in recall.
The latter option could also improve the running time, since multiple smaller
signatures may be faster to explore than a single large one.

@@

IsaCoSy is a tool for the Isabelle proof assistant, mostly written in Standard
ML. We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
with the 2015 version of Isabelle.

The following issues had to be overcome to make our benchmark applicable to
IsaCoSy:

\begin{itemize}
\item TIP includes a benchmark called \texttt{polyrec} whose types cannot be
  encoded in Isabelle. We strip out this type and the functions which depend on
  it before translating. It still appears in samples and contributes to the
  ground truth, which penalises IsaCoSy for being unable to explore such
  definitions.
\item When using a type in an IsaCoSy signature, that type's constructors will
  automatically be included in the exploration. Since those constructors will
  not appear in the ground truth (we use $\eta$-expanded wrappers instead, and
  even those may not be present in the current sample) this will unfairly reduce
  the calculated precision. To avoid this, we add a post-processing step which
  replaces all occurrences of a constructor with the corresponding wrapper, then
  discards any conjectures which involve functions other than those in the
  current sample. This presumably results in more work for IsaCoSy, exploring
  constructors unnecessarily, but it at least does not bring down the quality
  measures.
\item Since Isabelle is designed for theorem proving rather than programming, it
  requires every definition to be accompanied by proofs of exhaustiveness and
  termination. These are difficult to generate automatically, and don't exist in
  the case of partial functions like destructor wrappers. Hence we use the
  ``quick and dirty'' option in Isabelle, which lets us skip these proofs with
  the \texttt{sorry} keyword.
\item Partial functions cause problems during exploration, since they can throw
  an ``undefined'' exception which causes IsaCoSy to abort. We avoid this by
  pre-populating IsaCoSy's constraint set with these undefined expressions
  (for example $\Head(\Nil)$), hence preventing IsaCoSy from ever generating
  them.
\end{itemize}

\begin{figure}
  \centering
  \input{isacosytime.pgf}
  \caption{Running times of IsaCoSY on theories sampled from TIP. Each point
    is an individual run, with 31 per size except for 1 (14) and 2 (30). Colour
    indicates the total number of conjectures produced by each run, with failing
    processes (timeouts and out of memory) shown in red. Points are spread out
    horizontally to prevent overlaps. Box plots show median and quartiles, with
    whiskers showing points at 1.5$\times$~inter-quartile range.}
  \label{figure:isacosy_runtimes}
\end{figure}

@@

The most striking result is how rapidly IsaCoSy's running time increases with
sample size: from size 7 upwards, all runs failed (most timed out, a few aborted
early due to running out of memory). As we found with our preliminary testing,
this increase appears exponential, so even large increases to the timeout would
not produce many more successful observations. A significant amount of IsaCoSy's
time (around 50 seconds) was spent loading the generated theory (containing all
distinct datatypes and functions from TIP); this overhead is unfortunate, but
since it is constant across all sample sizes it does not affect our conclusions.

With so few successful datapoints it is difficult to make strong claims about
the precision/recall quality of IsaCoSy's output. It is clear from the quantity
of non-zero proportions that IsaCoSy is capable of discovering interesting
conjectures, and the graphs appear to follow the same shapes as those of
QuickSpec (decreasing precision, flat recall), although the error margins are
too wide to be definitive.

@@

\begin{table}
  \centering
  \begin{tabular}{ |r|l| }
    \hline
    \bfseries Sample Size & \bfseries Speedup
    \csvreader[]{speedups.csv}{}
    {\\\hline\csvcoli&\csvcolii} \\
    \hline
  \end{tabular}
  \caption{Speed up from using QuickSpec compared to IsaCoSy, as reported by
    Speedup-test for a confidence level of 95\%. Sizes 1 and 2 were not
    considered statistically significant due to having less than 30 samples.
    These are conservative estimates, since the 300 second time limit acts to
    reduce the measured time difference.}
  \label{table:speedups}
\end{table}

@@

{
  \centering
  \begin{tabular}{ |l|l|l| }
    \hline
                           & Found by IsaCoSy & Not found by IsaCoSy \\
    \hline
    Found by QuickSpec     & 28               & 19                   \\
    Not found by QuickSpec & 4                & 28                   \\
    \hline
  \end{tabular}
}

@@

{
  \centering
  \begin{tabular}{ |l|l|l| }
    \hline
              & Interesting & Uninteresting \\
    \hline
    IsaCoSy   & 32          & 137      \\
    QuickSpec & 47          & 301      \\
    \hline
  \end{tabular}
}

These totals were tested for independence using Boschloo's test, assuming a
binomial distribution, and gave an p-value of 0.111. With our (arbitrary) choice
of $\alpha = 0.05$, this is not significant enough to reject the null hypothesis
that their precision is equal.

Note that this is the only comparison where IsaCoSy performed better (although
not significantly so) with 19\% precision compared to QuickSpec's 14\%. However,
precision on its own is not the whole story: it can be increased at the expense
of recall by simply generating fewer conjectures. IsaCoSy's significantly lower
recall indicates that it may be overly-conservative in this way, compared to
QuickSpec.

@@

% We focus on the Tons of Inductive Problems (TIP) theorem proving
% benchmark~\cite{claessen2015tip}, since it has several desirable properties:

% All together, TIP provides 219 distinct function definitions and 343
%   theorem statements, which is enough to fully exercise current MTE tools.
% Benchmark problems include examples from the theorem proving literature
%   as well as common program verification tasks, ensuring the resulting corpus is
%   relevant to researchers and practitioners.
% TIP's maintainers provide a set of tools, including translators from TIP's
%   native format (based on SMT-Lib~\cite{BarFT-SMTLIB}) to Haskell and Isabelle
%   (which are used by MTE tools).

% To aid reproducibility, we use a cryptographic hash as our source of randomness:
% the $n$th sample of size $s$ uses the SHA256 hash of $n$, $s$ and the definition
% corpus; the latter prevents tailoring the corpus to influence the choices,
% although MTE tools could .

% To provide a more robust comparison of these two tools, we also applied a
% paired difference test: measuring, for each sampled theory, the difference in
% time, precision and recall between the two tools.

@@

Fundamentally, any mathematical reasoning system must decide on, and formalise,
what counts as ``the good'' in mathematics.  Obvious metrics such as ``true'' or
``provable'' include trivial tautologies, while at the same time failing to
capture the ``almost true'', which can be a valuable trigger for theory change,
as demonstrated by Lakatos in his case studies of mathematical
development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
proposed metric. Methods for finding out what that means have come from
neuro-scientists, such as the study by Zeki \etal{} in which they test whether
mathematicians' experiences of abstract beauty correlates with the same brain
activity as experiences of sensory beauty~\cite{zeki}; computer scientists, such
as Colton's metrics in table~\ref{table:colton} which are proposed based largely
on a computer scientist's ``intuition'' and argumentation about why a metric
would be important (and previous use of such metrics, and - in the case of
``surprisingingness'' - on a single quote from a mathematician\footnote{Similarly,
  classification of a MATHsAiD result as either interesting or
  uninteresting~\cite{roy} was made by the main system developer: in both cases
  the main developer had a background in mathematics.}; and mathematicians
themselves, such as Gowers' suggestion that we can identify features which are
commonly associated with good proofs~\cite{gowers}. All of these approaches rest
upon the assumption that it makes sense to speak of ``the good'' in mathematics.
This assumption is supported by concepts such as Erdos's ``The Book'' -- a
theoretical construct of a book which contains the most elegant proof of every
theorem (brought to life in~\cite{aigner2010proofs}). However, empirical studies
by psychologists such as Inglis call into question such assumptions: work by
Inglis and colleagues has shown that there is not a single standard of validity
among contemporary mathematicians %FIXME https://dspace.lboro.ac.uk/dspace-jspui/handle/2134/12150
nor beauty. %FIXME needs citation

The key to our benchmarking methodology is the simplifying assumption that
theorem proving problem sets are a good proxy for desirable MTE output. This
allows us to side-step this philosophical quagmire to produce concrete,
measurable values; it is also a core weakness, as there are compelling reasons
to refute it.

>>>>

% Derived from the template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{csvsimple}
\usepackage{float}
\restylefloat{table}
\usepackage{hhline}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows, calc, decorations.pathreplacing, positioning,
  shapes.geometric}
% etc.

@@

One quantitative approach which has gained traction is \emph{precision/recall
  analysis}. This has been applied to IsaCoSy, IsaScheme and HipSpec, the latter
providing a direct comparison of these three
tools~\cite{claessen2013automating}.\footnote{Unfortunately this comparison only
  reports prior results from the other tools, rather than reproducing them.
  This makes the numbers less comparable, since each tool was tested with
  slightly different definitions (e.g. whether or not the exponential and
  maximum functions were included).} To measure precision and recall we first
choose a set of definitions (such as those in figure \ref{figure:list_theory})
and a \emph{ground truth} to treat as the ``ideal'' set of conjectures we would
like a tool to produce for those definitions (e.g. this might include
equation~\ref{eq:mapreduce}). Each tool is run and their outputs are compared to
the ground truth. To score 100\% on precision and recall, all of the conjectures
which appear in the ground truth must be generated, and nothing else:

@@

We can summarise runtimes by choosing their median, as this is more robust than
the mean against long-running outliers, and hence represents performance for a
``typical'' input of that size. Unlike previous one-off evaluations, we can also
compute the \emph{spread} of our data, for example with the inter-quartile
range.

The precision and recall for multiple samples can be combined in two ways. We
can calculate their mean value, known as the \emph{average of the ratios}, to
find the expected precision and recall for runs of this size. Alternatively, we
can sum their numerators and denominators to get the \emph{ratio of the
  averages}: this normalises the output and ground truth cardinalities, allowing
precision and recall to be modelled as simple Bernoulli processes.

@@

We propose that running times are compared using a Wilcoxon signed-rank
test~\cite{wilcoxon1945individual}, since it does not require that the data be
normally distributed. In our application we follow a variant of the Speedup-Test
protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the differences
are our use of the paired form of the Wilcoxon test, and the fact that each run
was performed on a different sample (to reduce sampling bias, at the expense of
accuracy).

@@

We cannot use McNemar's test to compare precision, since each tool may generate
different conjectures. Instead, we use Boschloo's form of Barnard's test to
compare the proportions of wanted and unwanted conjectures generated by each
tool.

@@

We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
benchmark as our ground truth corpus~\cite{claessen2015tip}.

@@

Based on this we decided to benchmark sample sizes up to 20, since neither tool
seemed to perform well beyond that. The Speedup-Test protocol follows the
statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
so we pick 31 samples of each size in order to cross this (arbitrary)
threshold. This gave a total of 1240 runs. To keep the benchmarking time down to
a few days we chose a timeout of 300 seconds, since that covered most of the
successful QuickSpec and IsaCoSy results we saw, and longer times gave rapidly
diminishing returns. During analysis, duplicate samples (caused by our sampling
restrictions) were found and discarded, so only 14 samples of size 1 were used
and 30 of size 2.

@@

We chose TIP for our ground truth since it satisfies the desiderata of
section~\ref{section:prep}: each problem has standalone type and function
definitions, making their separation trivial; known examples from the software
verification and inductive theorem proving literature are included, ensuring
relevance to those fields; the format includes the higher-order functions and
inductive datatypes we are interested in; it is large enough to pose a challenge
to current MTE tools; plus it is accompanied by tooling to convert its custom
format (an extension of SMT-Lib~\cite{BarFT-SMTLIB}) into a variety of
languages, including Haskell and Isabelle.

@@

\begin{samepage}
\begin{verbatim}
(declare-datatypes
  (a)                       ;; Type parameter (element type)
  ((List                    ;; Type name
     (Nil)                  ;; Constructor (nullary)
     (Cons                  ;; Constructor (binary)
       (head a)             ;; Field name and type
       (tail (List a))))))  ;; Field name and type
\end{verbatim}
\end{samepage}

@@

\begin{samepage}
\begin{verbatim}
(define-fun
  (par (a)                   ;; Type parameter
    (constructor-Cons        ;; Function name
      ((x a) (xs (List a)))  ;; Argument names and types
      (List a)               ;; Return type
      (as                    ;; Type annotation
        (Cons x xs)          ;; Return value
        (List a)))))         ;; Return type

(define-fun
  (par (a)                        ;; Type parameter
    (destructor-head              ;; Function name
      ((xs (List a)))             ;; Argument name and type
      a                           ;; Return type
      (match xs                   ;; Pattern-match
        (case (Cons h t) h)))))   ;; Return relevant field
\end{verbatim}
\end{samepage}

@@

Another complication is TIP's ``native'' support for booleans and integers,
which allows numerals and symbols like \texttt{+} to appear without any
accompanying definition. To ensure consistency in the translations, we replace
all occurrences of such expressions with standard definitions written with the
``user-level'' \texttt{declare-datatypes} and \texttt{define-fun}
mechanisms.~\footnote{\texttt{Boolean} has \texttt{true} and \texttt{false}
  constructors; \texttt{Natural} has \texttt{zero} and \texttt{successor};
  \texttt{Integer} has unary \texttt{positive} and \texttt{negative}
  constructors taking \texttt{Natural}s, and a nullary \texttt{zero} for
  symmetry.}

@@

TIP comes with software to translate its definitions into Haskell and Isabelle
code, including comparison functions and random data generators suitable for
QuickCheck. We translate all 269 unique definitions into a single module/theory
which is imported on each run of the tools, although only those functions which
appear in the current sample are included in the signature and explored.

@@

\begin{sloppypar}
  \begin{itemize}
  \item We must decide what variables to include. We choose to add three
    variables for each type that appears as a function argument, except for
    types which have no QuickCheck data generators.
  \item We must \emph{monomorphise} all types. For example, functions like
    \texttt{constructor-Cons} are \emph{polymorphic}: they build lists of any
    element type, but we need to pick a specific type in order to know which
    random value generator to use. We resolve this (arbitrarily) by picking
    \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
      \texttt{*} (types); for kind \texttt{* -> *} (type constructors) we pick
      \texttt{[]} (Haskell's list type constructor). If these violate some
      type class constraint, we pick a suitable type non-deterministically from
      those in scope during compilation; if no suitable type is found, we give
      up and don't include that function.}
  \item Haskell functions are ``black boxes'', which QuickSpec can't compare
    during its exploration process. They are also curried, always taking one
    argument but potentially returning another function. QuickSpec lets us
    assign an arity to each function in the signature, from 0 to 5, so we pick
    the highest that is type-correct, since this avoids a proliferation of
    incomparable, partially-applied functions.
  \end{itemize}
\end{sloppypar}

\begin{figure}
  \centering
  \input{quickspectime.pgf}
  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
    is an individual run, with 31 per size except for 1 (14) and 2 (30). Colour
    indicates the total number of conjectures produced by each run, with failing
    processes shown in red. Points are spread out horizontally to prevent
    overlaps. Box plots show median and quartiles, with whiskers showing points
    at 1.5$\times$~inter-quartile range.}
  \label{figure:quickspec_runtimes}
\end{figure}

@@

The time taken to explore samples of different sizes is shown in
figure~\ref{figure:quickspec_runtimes}. Runs which timed out are shown in red,
and these occurred for all sample sizes but are more common for larger samples
(over half the runs for sample size 20 timed out). Runs which succeeded mostly
finished within 30 seconds, generating few or no conjectures; those taking
longer generated more conjectures, with the most being 116 conjectures from a
sample of size 16.

Precision and recall are shown in figure~\ref{figure:quickspec_precRec}, which
we model as simple Bernoulli processes: the generated conjectures (wanted or
not) are like coin tosses whose bias is the precision; likewise, the ground
truth theorems (found or not) are like coin tosses biased by the recall. Note
that this simplification does not explain how many conjectures the tool decides
to output for each run.

QuickSpec's precision gets worse as the sample size grows. Since QuickSpec
generates monotonically more conjectures as definitions are added to a signature
(assuming sufficient testing), it can't be finding fewer correct conjectures at
larger sizes. This is supported by the relative flatness of the recall results.
Rather, the number of conjectures generated is increasing at a higher rate than
the size of the ground truth. These extra conjectures may involve those
``padding'' definitions in a sample which don't contribute to its ground truth,
or may be ``uninteresting'' relationships between the dependencies of different
ground truth theorems.

This indicates two potential improvements to the QuickSpec algorithm (as far as
this benchmark is concerned). The deluge of generated conjectures could be
filtered down to a more desirable sub-set by another post-processing step.
Alternatively, rather than exploring all of the given definitions together,
multiple smaller signatures could be selected from the input by predicting which
combinations are likely to lead to interesting conjectures; this would avoid
both the ``padding'' and the cross-dependency relationships. Both of these
methods could improve the precision, although care would be needed to avoid a
large reduction in recall. The latter option could also improve the running
time, since (based on figure~\ref{figure:quickspec_runtimes}) multiple smaller
signatures may be faster to explore than a single large one. Such improvements
would also need to be ``generic'', to avoid over-fitting to this particular
benchmark.

@@

IsaCoSy is a tool for the Isabelle proof assistant, mostly written in Standard
ML. We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
with the 2015 version of Isabelle. The following issues had to be overcome to
make our benchmark applicable to IsaCoSy:

\begin{itemize}
\item TIP includes a benchmark called \texttt{polyrec} whose types cannot be
  encoded in Isabelle. We strip out this type and the functions which depend on
  it before translating. It still appears in samples and contributes to the
  ground truth, which penalises IsaCoSy for being unable to explore such
  definitions.
\item When using a type in an IsaCoSy signature, that type's constructors will
  automatically be included in the exploration. Since those constructors will
  not appear in the ground truth (we use $\eta$-expanded wrappers instead, and
  even those may not be present in the current sample) this will unfairly reduce
  the calculated precision. To avoid this, we add a post-processing step which
  replaces all occurrences of a constructor with the corresponding wrapper, then
  discards any conjectures which involve functions other than those in the
  current sample. This presumably results in more work for IsaCoSy, exploring
  constructors unnecessarily, but it at least does not bring down the quality
  measures.
\item Since Isabelle is designed for theorem proving rather than programming, it
  requires every definition to be accompanied by proofs of exhaustiveness and
  termination. These are difficult to generate automatically, and don't exist in
  the case of partial functions like destructor wrappers. Hence we use the
  ``quick and dirty'' option in Isabelle, which lets us skip these proofs with
  the \texttt{sorry} keyword.
\item Partial functions cause problems during exploration, since they can throw
  an ``undefined'' exception which causes IsaCoSy to abort. We avoid this by
  pre-populating IsaCoSy's constraint set with these undefined expressions
  (for example \texttt{(destructor-head constructor-Nil)}), hence preventing
  IsaCoSy from ever generating them.
\end{itemize}

\begin{figure}
  \centering
  \input{isacosytime.pgf}
  \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
    is an individual run, with 31 per size except for 1 (14) and 2 (30). Colour
    indicates the total number of conjectures produced by each run, with failing
    processes (timeouts and out of memory) shown in red. Points are spread out
    horizontally to prevent overlaps. Box plots show median and quartiles, with
    whiskers showing points at 1.5$\times$~inter-quartile range.}
  \label{figure:isacosy_runtimes}
\end{figure}

@@

The most striking result is how rapidly IsaCoSy's running time, shown in
figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
size 6 failed, with most timing out and a few aborting early due to running out
of memory. Like in our preliminary testing, this increase appears exponential,
so even large increases to the timeout would not produce many more successful
observations. A significant amount of IsaCoSy's time (around 50 seconds) was
spent loading the generated theory (containing all distinct datatypes and
functions from TIP); this overhead is unfortunate, but since it is constant
across all sample sizes it does not affect our conclusions.

With so few successful datapoints it is difficult to make strong claims about
the precision/recall quality of IsaCoSy's output, shown in
figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
proportions that IsaCoSy is capable of discovering interesting conjectures, and
the graphs appear to follow the same shapes as those of QuickSpec (decreasing
precision, flat recall), although the error margins are too wide to be
definitive.

@@

\begin{table}
  \centering
  \begin{tabular}{ |r|l| }
    \hline
    \bfseries Sample Size & \bfseries Speedup
    \csvreader[]{speedups.csv}{}
    {\\\hline\csvcoli&\csvcolii} \\
    \hline
  \end{tabular}
  \caption{Speed up from using QuickSpec compared to IsaCoSy, as reported by
    Speedup-Test for a confidence level of 95\%. Sizes 1 and 2 were not
    considered statistically significant due to having less than 30 samples.
    These are conservative estimates, since the 300 second time limit acts to
    reduce the measured time difference.}
  \label{table:speedups}
\end{table}

@@

\begin{table}[H]
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
    \hline
    \bfseries Found by QuickSpec     & 28 & 19 \\
    \bfseries Not found by QuickSpec & 4  & 28 \\
    \hline
  \end{tabular}
\end{table}

@@

\begin{table}[H]
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
              & \bfseries Interesting & \bfseries Uninteresting \\
    \hline
    \bfseries IsaCoSy   & 32          & 137      \\
    \bfseries QuickSpec & 47          & 301      \\
    \hline
  \end{tabular}
\end{table}

These totals were tested for independence using Boschloo's
test~\cite{lydersen2009recommended}, assuming a binomial distribution and
conditioning on the number of generated conjectures (as if this were fixed by
design). This produces a p-value of 0.111, which given our (arbitrary) choice of
$\alpha = 0.05$ is not significant enough to reject the null hypothesis that
their precision is equal.

Note that this is the only comparison where IsaCoSy performed better (although
not significantly so) with 19\% precision compared to QuickSpec's 14\%. However,
precision on its own is not the whole story: it can be increased at the expense
of recall by making the tool more conservative. IsaCoSy's significantly lower
recall indicates that it may be overly-conservative compared to QuickSpec, which
would also explain its precision.

@@

@@

Fundamentally, any mathematical reasoning system must decide on, and formalise,
what counts as ``the good'' in mathematics.  Obvious metrics such as ``true'' or
``provable'' include trivial tautologies, while at the same time failing to
capture the ``almost true'', which can be a valuable trigger for theory change,
as demonstrated by Lakatos in his case studies of mathematical
development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
proposed metric. Neuro-scientists such as Zeki \etal{} have attempted to shed
light on this by testing whether mathematicians' experiences of abstract beauty
correlates with the same brain activity as experiences of sensory
beauty~\cite{10.3389/fnhum.2014.00068}. Metrics such as those in
table~\ref{table:colton}, from computer scientists like Colton, are based
largely on ``intuition'', plausibility arguments about why a metric would be
important, and previous use of such metrics (in the case of
``surprisingingness'', from a single quote from a mathematician).
Opinions from mathematicians themselves include Gowers' suggestion that we can
identify features which are commonly associated with good
proofs~\cite{gowers2000two}, Erdos's famous idea of ``The Book'' (a theoretical
book containing the most elegant proof of every theorem, brought to life
in~\cite{aigner2010proofs}) as well as McCasland's evaluation of the
interestingness of MATHsAiD's output~\cite{roy} (of which he was the main system
developer).

All of these approaches rest upon the assumption that it even makes sense to
speak of ``the good'' in mathematics. However, empirical psychological studies
call into question such assumptions: for example, work by Inglis and colleagues
has shown that there is not even a single standard of \emph{validity} among
contemporary mathematicians~\cite{inglis2013mathematicians}.

Whilst these difficulties are real and important, we cannot ignore the fact that
mathematics is nevertheless being practised around the world; and similarly that
researchers have forged ahead to develop a variety of tools for automated
exploratory mathematics. If we wish to see these tools head in a useful,
fruitful direction then \emph{some} method is needed to compare their
approximate ``quality'' in a concrete, measurable way.
